An Introduction to Rocker: Docker Containers for R

### 2017-09-13 manuscript.Rmd
update citation format

The `debian:testing` stack aims to makes the most efficient use of
upstream builds -- the precompiled `.deb` binaries provided by the
Debian repositories.  It is both quicker and easier to install software
from binaries, since the package manager (`apt`) manages the necessary
(binary) dependencies and bypasses the time-consuming process of compiling
from source.  Basing this stack on `debian:testing` means that much more
recent versions of commonly-used libraries and compilers are available
as binaries than would be found in a Debian stable release.  In order
to provide access to the most recent avialble binaries, this stack uses
apt-pinning \citep{apt-pinning} to allow the `apt` package manager to also
install binaries from `debian:unstable`, which represents the most recent,
bleeding edge of packages built for Debian when necessary.  For instance,
the `r-base` image provided by Rocker installs the most recent version of
R as a binary from Debian `unstable`.  Similarly, recent versions of many
popular R packages can also be installed through the package manager,
e.g. `apt-get install r-cran-xml`.  This can be particularly helpful
for packages with external system dependencies (such as `libxml2-dev`
in this example) which cannot be installed from the R console as they 
are system dependencies rather than R packages installed from within R.
We should note, however, that only about 500 of the currently almost
11000 CRAN packages are available as Debian packages.

@@

1. These Dockerfiles require occassional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt-pinning}.
4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team. [@debian-security]
<!-- DE: Are you sure about that point 4 ? -->
<!-- CB: I dunno, just quoting from https://www.debian.org/releases/testing/ -->

@@

Several of the images in this stack are oriented towards the R
development community. `r-devel`, `drd`, `r-devel-san` &
`r-devel-ubsan-clang` all add a copy of the development version of R
side-by-side the current release of R provided by `r-base`.  On these
images, the development version is aliased to `RD` to distinguish from
the current release, `R`.  As the names suggest, each provide slightly
different configurations.  Of particular interest are the images
providing development R built with support for C/C++ address and
undefined-behavior sanitizers, which are somewhat more difficult to
configure \citep{nedd_sanitizers}.

@@

**MRAN archives**: To facilitate installation of only contemporaneous
versions of R packages on these images, the default CRAN mirror from which
to install R packages is fixed to a snapshot of CRAN corresponding to
the last date for which that version of R was the latest release. These
snapshots are provided by the MRAN archive created by Revolution
Analytics (now part of Microsoft). It archives daily snapshots
of all of CRAN from which a user can install packages with the usual
`install.packages()` function  [@MRAN]. Users can always override this
default by passing any current CRAN repository explicitly.  Unlike CRAN,
Bioconductor only updates its repositories through bi-annual releases
aligned to R's spring release schedule. Thus, Bioconductor packages
can be installed in the usual way using `bioclite`, which automatically
selects the Bioconductor release corresponding to the version of R in use.

**Version tags**:  The version tags are propegated throughout this
stack: _e.g._, `rocker/tidyverse:devel` will provide the currently
released versions of the R packages in the `tidyverse` [@tidyverse]
installed on the nightly build of R-devel.  Developers building packages
on this stack are encouraged to tag their images accordingly as well.
Table 3 indicates which versions of R are currently available in the
stack, going back to `3.1.0`.  While older versions may be added to
the stack at a later date, we note that the MRAN snapshots began in
starting 2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
be built from a separate Dockerfile, enabling minor differences in the
build instructions to accomodate changing dependencies.  Dockerfiles for
past versions (e.g. < `3.4.0` currently) are intended to remain static
over the long term, while the tag for the current version, `latest`,
and `devel` may be tweaked to accomodate new features or dependencies.

@@

The `rstudio` image includes a lightweight, easy-to-use and
docker-friendly `init` system, s6 [@s6] for running persistant services,
including the RStudio server.  This system provides a convenient way for
downstream Dockerfile developers to add additional persistent services
(such as an `ssh` server) to a single container, or additional start-up
or shutdown scripts that should be run when an instance starts up or
shuts down.  The `rstudio` image uses such a start-up script to configure
user settings such as login password and permissions through environmental
variables at run time.  <!-- link wiki howto page? -->

@@

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R envionment to a remote server, such
as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
Jetstream servers [@jetstream], or private or institutional server hardware.  Being
able to test code in a predictable, pre-configured R environment on a
local machine and to then run the same code in an identical environment on a
remote server (e.g. for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such contanerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

@@

<!-- comment on use in instructional context -->
This portablility is similarly valuable in an instructional context. Requiring students
to install all necessary software on personal laptops can be particularly challenging
for short workshops, where download and installation time and troubleshooting across
heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
By deploying a Rocker image or Rocker-derived image (see extensibility) on a cloud machine,
an instructor can easily provide all students access to the pre-configured software environment
using only the browser on their laptops.  Unlike a vendor-specific virtual machine image (e.g. Amazon Machine Images, AMIs),
this approach does not preclude students being able to run the same software locally later on.
After installing Docker for their machine they too can run the course-specific software environment
easily on their own laptops and deploy it on other cloud providers.  This strategy has proven
effective in our own experience in both workshops and semester-length courses. Similar Docker-based
cloud deployments have been scaled to courses of 100s of students, e.g. at Duke [@Mine] and
UC Berkeley [@data8].

@@

<!-- Blah blah blah ... link to documentation instead(?) -->
**Root vs user instances**: By default, all rocker containers run as
root. This is consistent with standard Docker practice, allowing both
interactive users and downstream Dockerfiles to install software directly
without having to switch to root.  The main snag of this approach comes if
a user links a local volume and modifies files on that volume, which will
result in the local file being owned by `root` and not the default user.
To avoid this, specify the non-root user at runtime (username `docker`
on the `r-base` stack, or `rstudio` on images deriving from rstudio)
when running an interactive shell.  For users accessing R through RStudio,
this is not necessary.  The docker container will run as root, but a user
logs in through the RStudio server web interface as the non-root user
"rstudio", and thus any changes made to linked volumes will not alter file
permissions on the home directory.  For more details on user permissions,
see the Rocker wiki. [@rockerwiki]

@@

This sandboxing feature of Docker containers is also valuable in the 
remote computing context, where it can allow system administrators to
provide users with administrative (root) privilages to alter and install
software within a container without altering the computational environment
of other users on the remote machine.  Unlike traditional virtual machines,
these containers do not impose a large footprint of reserved resources --
a typical host can easily support 100s of containers [@docker-vs-vm].

@@

[@Boettiger2015]

@@

<!-- see docs instead -->
Rocker images provide a few utility functions to streamline this process, including the 
`littler` [@littler] scripts which provide a concise syntax for installing packages in Dockerfiles, e.g.

@@

Users writing their own Dockerfiles are encouraged to follow the same practices as the
Rocker Project, such as the Dockerfile Best Practices [@best-practices], the use of automated builds,
and when appropriate, versioned tags.  Users can ensure their automated builds are rebuilt
every time the relevant upstream DOckerfile is updated by using build triggers on Docker Hub.

>>>>

The `debian:testing` stack aims to makes the most efficient use of
upstream builds -- the precompiled `.deb` binaries provided by the
Debian repositories.  It is both quicker and easier to install software
from binaries, since the package manager (`apt`) manages the necessary
(binary) dependencies and bypasses the time-consuming process of compiling
from source.  Basing this stack on `debian:testing` means that much more
recent versions of commonly-used libraries and compilers are available
as binaries than would be found in a Debian stable release.  In order
to provide access to the most recent avialble binaries, this stack uses
apt-pinning \citep{apt_pinning} to allow the `apt` package manager to also
install binaries from `debian:unstable`, which represents the most recent,
bleeding edge of packages built for Debian when necessary.  For instance,
the `r-base` image provided by Rocker installs the most recent version of
R as a binary from Debian `unstable`.  Similarly, recent versions of many
popular R packages can also be installed through the package manager,
e.g. `apt-get install r-cran-xml`.  This can be particularly helpful
for packages with external system dependencies (such as `libxml2-dev`
in this example) which cannot be installed from the R console as they 
are system dependencies rather than R packages installed from within R.
We should note, however, that only about 500 of the currently almost
11000 CRAN packages are available as Debian packages.

@@

1. These Dockerfiles require occassional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt_pinning}.
4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team \citep{debian_security}.
<!-- DE: Are you sure about that point 4 ? -->
<!-- CB: I dunno, just quoting from https://www.debian.org/releases/testing/ -->

@@

Several of the images in this stack are oriented towards the R
development community. `r-devel`, `drd`, `r-devel-san` &
`r-devel-ubsan-clang` all add a copy of the development version of R
side-by-side the current release of R provided by `r-base`.  On these
images, the development version is aliased to `RD` to distinguish from
the current release, `R`.  As the names suggest, each provide slightly
different configurations.  Of particular interest are the images
providing development R built with support for C/C++ address and
undefined-behavior sanitizers, which are somewhat more difficult to
configure \citep{edd_sanitizers}.

@@

**MRAN archives**: To facilitate installation of only contemporaneous
versions of R packages on these images, the default CRAN mirror from which
to install R packages is fixed to a snapshot of CRAN corresponding to
the last date for which that version of R was the latest release. These
snapshots are provided by the MRAN archive created by Revolution
Analytics (now part of Microsoft). It archives daily snapshots
of all of CRAN from which a user can install packages with the usual
`install.packages()` function  \citep{MRAN}. Users can always override this
default by passing any current CRAN repository explicitly.  Unlike CRAN,
Bioconductor only updates its repositories through bi-annual releases
aligned to R's spring release schedule. Thus, Bioconductor packages
can be installed in the usual way using `bioclite`, which automatically
selects the Bioconductor release corresponding to the version of R in use.

**Version tags**:  The version tags are propegated throughout this
stack: _e.g._, `rocker/tidyverse:devel` will provide the currently
released versions of the R packages in the `tidyverse` \citep{tidyverse}
installed on the nightly build of R-devel.  Developers building packages
on this stack are encouraged to tag their images accordingly as well.
Table 3 indicates which versions of R are currently available in the
stack, going back to `3.1.0`.  While older versions may be added to
the stack at a later date, we note that the MRAN snapshots began in
starting 2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
be built from a separate Dockerfile, enabling minor differences in the
build instructions to accomodate changing dependencies.  Dockerfiles for
past versions (e.g. < `3.4.0` currently) are intended to remain static
over the long term, while the tag for the current version, `latest`,
and `devel` may be tweaked to accomodate new features or dependencies.

@@

The `rstudio` image includes a lightweight, easy-to-use and
docker-friendly `init` system, s6 \citep{s6} for running persistant services,
including the RStudio server.  This system provides a convenient way for
downstream Dockerfile developers to add additional persistent services
(such as an `ssh` server) to a single container, or additional start-up
or shutdown scripts that should be run when an instance starts up or
shuts down.  The `rstudio` image uses such a start-up script to configure
user settings such as login password and permissions through environmental
variables at run time.  <!-- link wiki howto page? -->

@@

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R envionment to a remote server, such
as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
Jetstream servers \citep{jetstream}, or private or institutional server hardware.  Being
able to test code in a predictable, pre-configured R environment on a
local machine and to then run the same code in an identical environment on a
remote server (e.g. for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such contanerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

@@

<!-- comment on use in instructional context -->
This portablility is similarly valuable in an instructional context. Requiring students
to install all necessary software on personal laptops can be particularly challenging
for short workshops, where download and installation time and troubleshooting across
heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
By deploying a Rocker image or Rocker-derived image (see extensibility) on a cloud machine,
an instructor can easily provide all students access to the pre-configured software environment
using only the browser on their laptops.  Unlike a vendor-specific virtual machine image (e.g. Amazon Machine Images, AMIs),
this approach does not preclude students being able to run the same software locally later on.
After installing Docker for their machine they too can run the course-specific software environment
easily on their own laptops and deploy it on other cloud providers.  This strategy has proven
effective in our own experience in both workshops and semester-length courses. Similar Docker-based
cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
UC Berkeley \citep{data8}.

@@

<!-- Blah blah blah ... link to documentation instead(?) -->
**Root vs user instances**: By default, all rocker containers run as
root. This is consistent with standard Docker practice, allowing both
interactive users and downstream Dockerfiles to install software directly
without having to switch to root.  The main snag of this approach comes if
a user links a local volume and modifies files on that volume, which will
result in the local file being owned by `root` and not the default user.
To avoid this, specify the non-root user at runtime (username `docker`
on the `r-base` stack, or `rstudio` on images deriving from rstudio)
when running an interactive shell.  For users accessing R through RStudio,
this is not necessary.  The docker container will run as root, but a user
logs in through the RStudio server web interface as the non-root user
"rstudio", and thus any changes made to linked volumes will not alter file
permissions on the home directory.  For more details on user permissions,
see the Rocker website, <https://rocker-project.org>.

@@

This sandboxing feature of Docker containers is also valuable in the 
remote computing context, where it can allow system administrators to
provide users with administrative (root) privilages to alter and install
software within a container without altering the computational environment
of other users on the remote machine.  Unlike traditional virtual machines,
these containers do not impose a large footprint of reserved resources --
a typical host can easily support 100s of containers \cite{what-docker}.

@@

\citep{@Boettiger2015}

@@

<!-- see docs instead -->
Rocker images provide a few utility functions to streamline this process, including the 
`littler` \citep{littler} scripts which provide a concise syntax for installing packages in Dockerfiles, e.g.

@@

Users writing their own Dockerfiles are encouraged to follow the same practices as the
Rocker Project, such as the Dockerfile Best Practices \citep{best_practices}, the use of automated builds,
and when appropriate, versioned tags.  Users can ensure their automated builds are rebuilt
every time the relevant upstream Dockerfile is updated by using build triggers on Docker Hub.
